---
title: "Country-specific analysis of individual-level data"
author: "Dan Weinberger"
date: "May 13, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Install development version from GitHub. This requires that the 'devtools'
# package be installed.
if (!require("devtools")) {
  install.packages("devtools")
}
devtools::install_github("weinbergerlab/ExcessILI", upgrade =  F)
library(ExcessILI)
library(ggplot2)
library(reshape2)
library(dplyr)
library(tidyr)
```

## Import data
These datasets contain the individual-level mortality records for each country. The files were initially stored as Microsoft Excel databases; however the files were quite large and unwieldy, so they have been converted to the R data format (.rds), which stores the data more efficiently.

*The last 2 letters on the file name indicate the country*

If you are using a PC, save the .rds file for your country in the same directory as this .Rmd file. If you are using a mac, you will need to change the directory below.
```{r cars}
d1 <- readRDS('./Data/IndividualICD10_mx.rds')

```

Format the date variable
```{r}
d1$date <- as.Date( paste0(d1$Year,'-',d1$Month,'-','01' )) #Format the date variable
```

##Look at structure of the data
```{r}
head(d1)
```

Get the variable names of the data

```{r}
names(d1)
```

Format the date variable
```{r}
d1$date<- as.Date( paste0(d1$Year,'-',d1$Month,'-','01' )) #Format the date variable
```



### Then do some basic explorations. 

What is the distribution of Age *groups*
```{r hist2}
hist(d1$Age.group)
```

What is the distribution of dates?
```{r hist3}
hist(d1$date, breaks=10)
```

### Which codes are the most commonly used in this database?
Make a table of the codes, sorted most to least common. Try to modify this and look at the distribution of secondary DX codes (e.g., DX2,)

```{r freq.codes, echo=FALSE}
tab1 <- table(d1$dx1)

sort(tab1, decreasing=T)

```

## And let's make a code for each of the ICD10 chapters to see bigger-picture patterns
```{r}
chapter.prim<-substr(d1$dx1,1,1)
```

**Remember** if you ever want to see the full data frame, just type "View(DATAFRAMENAME) so in this case, View(d1)



Frequency of each ICD10 chapter
```{r}
tab2 <- table(chapter.prim)

sort(tab2, decreasing = T)
```

## Let's look at deaths due to pneumonia

Extract first 3 digits from DX1 ICD10 code and save as a vector called icd10.3digits
```{r}
d1$icd10.3digits<-substr(d1$dx1,1,3) #extract 1st 3 digits from IC10 code
```

view the first 10 entries
```{r}
d1$icd10.3digits[1:10] 
```

Now we are going to create a variable to flag all deaths that have a code between J12 and J1
```{r}
d1$j12_j18<-rep(0, nrow(d1)) #Initialize variable with 0s
```

Replace with 1 if the 3 digit ICD10 dx1 is J12,J13,J14,J15,J16,J17,J18 (note there are more efficient ways to code this using greater than and less than symbols)

```{r}
d1$j12_j18[d1$icd10.3digits %in% c('J12', 'J13', 'J14', 'J15', 'J16', 'J17', 'J18') ] <- 1
```

## Let's aggregate now by date
We sum the 0/1 variable we created for J12_J18 by date. this counts the number of cases per time period. Here we are using 'pipes'
```{r}


d2 <- d1 %>% #creating a new data frame called 'd2', input data frame d1
    group_by(date) %>%    #group by date
    summarize('j12_j18' = sum(j12_j18) )  #take the sum of j12_18 across all values of the data


```

## Plot your time series

First, a basic, ugly plot

```{r}

 p1 <- ggplot(d2, aes(x=date, y=j12_j18)) + #define the data frame and x and y variables
      geom_line()   #say that you want a line plot
 
 p1
```
Let's make it look nicer...

```{r}

 p2 <- ggplot(d2, aes(x=date, y=j12_j18)) +
      geom_line(col='blue')  +
      ylab("Number of pneumonia deaths") +  #add y axis label
      xlab("Date") +   #add x axis label
      theme_classic() +  #choose a 'theme'
      geom_vline(xintercept=as.Date('2006-01-01'), col='gray', lty=2) #add reference line for vaccine introduction
 
 p2
```

##Now let's stratify by age group data
0 if <2 mo
1 if 2-11 mo
2 if 12-23 mo
3 if 24-59 mo

Generate some labels that can be used later for plotting
```{r}
#Create labels we can use 
age.labels<-c('<2m','2-11m','12-23m','24-59m')
```

Aggregate variable J12_18 by age group and date
```{r}

d3 <- d1 %>% #creating a new data frame called 'd3', input data frame d1
    group_by(Age.group, date) %>%    #group by date
    summarize('j12_j18' = sum(j12_j18) ) %>% #take the sum of j12_18 across all values of the data
    ungroup()
```

Change variable names
```{r}
names(d3) <- c('agec','date','J12_18')
```

Add labels to age categories after converting to a factor

```{r}
d3$agec <- as.factor(d3$agec)

d3$agec <- factor(d3$agec, levels=c('0','1','2','3'), 
                  labels=c('<2m','2-11m','12-23m','24-59m'))
```



sort the rows of the data frame by age and date
```{r}
d3 <- d3 %>%
  arrange(agec, date) #sort by agec then by date
```

## Plot by group

Previously we made a single time series plot. Now we want to make a plot for each age group. We can do this with a 'facet wrap". We will modify the code that we wrote earlier. 

```{r, fig.width=6, fig.height=6}

#These first lines of code are identical to what we had earlier
p3 <- ggplot(d3, aes(x=date, y=J12_18)) +
      geom_line(col='blue')  +
      ylab("Number of pneumonia deaths") +  #add y axis label
      xlab("Date") +   #add x axis label
      theme_classic() +  #choose a 'theme'
      geom_vline(xintercept=as.Date('2006-01-01'), col='gray', lty=2) + #add reference line for vaccine introduction
 
  #These last few lines
      facet_wrap(~agec, scales='free_y') + #different plot for each age group, each with unique scale (scale='free')
      theme(panel.spacing= unit(2,'lines') , axis.text.x=element_text(angle=90))
  
p3 #print the plot
```


For future use, we could save this code as a generic function, which we could then call easily with different variable names and groups

Define the function 'ts.plot.fun'. Note that this looks very similar to the code above, except we are subbing in the real variable names with placeholders. 
```{r}


ts.plot.fun <- function(ds, time.var=date, yvar,multiple_group=F, group.var, vax.intro.date ){
  p1 <- ggplot(ds, aes_string(x=time.var, y=yvar)) +
      geom_line() +
      ylab("Number of pneumonia deaths") +
      xlab("Date") +
      theme_classic() +
      geom_vline(xintercept=vax.intro.date, col='gray', lty=2) +
  
        facet_wrap(group.var , scales='free_y') +
        theme(panel.spacing= unit(2,'lines') , axis.text.x=element_text(angle=90)) 

  return(p1)
}
  
```

Call the function
```{r, fig.width=6, fig.height=6}
  ts.plot.fun(ds=d3,
              time.var='date', 
              yvar='J12_18', 
              multiple_group = T, 
              group.var='agec',
              vax.intro.date = as.Date('2006-01-01'))

```




## Now we need to extract some control variables
1) Flag any record that has a potential pneumococcal-related code anywhere in the record

dx.all <- d1[,c("dx1","dx2","dx3","dx4","dx5","dx6")] 
Creates a cut of data that just has the DX codes

```{r}
dx.all <- d1[,c("dx1","dx2","dx3")] 
```

We are interested in codes in the range between J12, J18
```{r}
J12_code_range <- c("J12", "J13", "J14", "J15", "J16", "J17", "J18")
```

For each observation (row), search for the J12-J18 codes in the first 3 digits of each of the DX codes
```{r}
d1$J12_J18_any <-  1 * apply(dx.all,
                             1, #Repeat by row/observation
                             function(x) {
                               max(J12_code_range %in%
                                 substr(x, 1, 3)
                               )
                             })
```

Then create a variable that flags codes that are possibly related to pneumococcus
Here are some codes that might be pneumococcal-related
```{r}
other.related.dx <-
  c("A40","A49","B953","R652","H10","H65", 
    "H66","G00","G01","G02","G03","G04","J12", "J13",
    "J14", "J15", "J16", "J17", "J18" )
```

then search row by row again for matches ot these codes
```{r}
d1$possible_pneumo_code <-  1 * apply(dx.all,
                             1, #Repeat by row/observation
                             function(x) {
                               max(other.related.dx %in%
                                 substr(x, 1, 3)
                               )
                             })
```

And let's create another control variable that consists of the A00-B99 chapter. We will flag a record IF it has codes in the range A00-B99 and does NOT have any pneumococcal-related codes OR any rotavirus-related codes. This is important because rotavirus vaccine was introduced around the same time, and rotavirus diarrhea declined. If we included that in the control, it could mask the benefits of PCVs because the comparison and the disease of interest of both declining, but for different reasons

A00-B99 primary cause of death:
Search for the letter A or B in DX1, and set to 1 if it is there
```{r}

d1$A00_B99_prim <-  1 * apply(dx.all[,'dx1', drop=F],
                             1, #Repeat by row/observation
                             function(x) {
                               max(c('A','B') %in%
                                 substr(x, 1, 1)
                               )
                             })

#Set=0 if A00-A09
d1$A00_B99_prim[substr(d1$dx1, 1, 2) =="A0"] <- 0

#Set=0 if possible pneumococcal code
d1$A00_B99_prim[d1$possible_pneumo_code==1] <- 0
```

Check your work. We should expect to see that these two categories are mutually exclusive (shouldn't have possible pneumo code=1 and A00_B99_prim==1)
```{r}
table(d1$A00_B99_prim, d1$possible_pneumo_code) # Good.
```

Also extract some more specific sub-chapter codes
```{r}
#flags Dx1 if it starts with A1
d1$A15_A19_prim <-  1 * apply(dx.all[,'dx1', drop=F],
                             1, #Repeat by row/observation
                             function(x) {
                               max(c('A1') %in%
                                 substr(x, 1, 2)
                               )
                             })

#Set=0 if possible pneumococcal code
d1$A15_A19_prim[d1$possible_pneumo_code==1] <- 0

```
Check your work
```{r}
table(d1$A15_A19_prim, d1$possible_pneumo_code)
```

And let's add in a few more controls. for each of these, we are using broad ICD10 chapters based on the starting letter For example, all primary codes that start with P or Q, or R. And for each of these we exclude the death if a potential pneumococcal code is anywhere on the record.
```{r}
d1$p_prim<- 1*(substr(d1$dx1,1,1)=='P') #logical test for whether code starts with P
d1$p_prim[d1$possible_pneumo_code==1]<-0 #don't count of pneumococcal code is present

d1$q_prim<- 1*(substr(d1$dx1,1,1)=='Q') #logical test for whether code starts with P
d1$q_prim[d1$possible_pneumo_code==1]<-0 #don't count of pneumococcal code is present

d1$r_prim<- 1*(substr(d1$dx1,1,1)=='R') #logical test for whether code starts with P
d1$r_prim[d1$possible_pneumo_code==1]<-0 #don't count of pneumococcal code is present

d1$w_prim<- 1*(substr(d1$dx1,1,1)=='W') #logical test for whether code starts with P
d1$w_prim[d1$possible_pneumo_code==1]<-0 #don't count of pneumococcal code is present

d1$g_prim<- 1*(substr(d1$dx1,1,1)=='G') #logical test for whether code starts with P
d1$g_prim[d1$possible_pneumo_code==1]<-0 #don't count of pneumococcal code is present

```

## Now aggregate the outcome and the controls

We could do this similar to how we did it previously. 
```{r}

e2 <- d1 %>%
  group_by(Age.group, date) %>%
  summarize(j12_j18 = sum(j12_j18) ,
            A15_A19_prim = sum(A15_A19_prim),
            A00_B99_prim = sum(A00_B99_prim),
            p_prim = sum(p_prim),
            q_prim = sum(q_prim),
            r_prim = sum(r_prim),
            w_prim = sum(w_prim),
            g_prim = sum(g_prim),
            all.visits = n()
            ) %>%
  ungroup() 

names(e2)[1:2] <- c('agec','date')
  
```

Check the structure of the data
```{r}
str(e2)
```

**Important note **
It is critical for the analysis that every time period for every stratum is represented. If there were no case/deaths that occur for a particular group in a given time period, a 0 should be included. One way to do this is wih the complete() function. for example if we want to make sure j12_18 and A15_19 columns have a row for each age group and date combination:

```{r}

e2 <-  e2   %>%
      group_by(agec) %>%
      complete(date=seq.Date( from=min(e2$date, na.rm=T), to=max(e2$date, na.rm=T), by='month')) %>%
      replace_na( list(j12_j18=0,                   
                       A15_A19_prim=0 ,                  
                       A00_B99_prim=0,                  
                       p_prim=0,                  
                       q_prim=0,                  
                       r_prim=0,                  
                       w_prim=0,                  
                       g_prim=0,                  
                       all.visits=0)) %>%
  ungroup()

```
## Just about ready to start some analyses

We just need to do some data transformations. To do this, we will use some pipes...we will group by age, then log transform and scale the variables(mean=0, sd=1)


```{r}

#define a function to scale the vairables
scale_fun <- function(x){
  (x - mean(x, na.rm=TRUE)) / sd(x, na.rm=TRUE)
}

e2 <- e2 %>%
  group_by(agec) %>%
  mutate(   log_A15_A19_prim = scale_fun(log(A15_A19_prim+0.5)),
            log_A00_B99_prim = scale_fun(log(A00_B99_prim+0.5)),
            log_p_prim = scale_fun(log(p_prim+0.5)),
            log_q_prim = scale_fun(log(q_prim+0.5)),
            log_r_prim = scale_fun(log(r_prim+0.5)),
            log_w_prim = scale_fun(log(w_prim+0.5)),
            log_g_prim = scale_fun(log(g_prim+0.5)),
            log_all.visits = scale_fun(log(all.visits+0.5))
  ) %>%
  ungroup()

```




Reshape the dataset from 'wide' to 'long'
```{r}

#identify the columns named 'log'

log_var_names <- names(e2)[grep('log_',names(e2))]  #grep searches for the string 'log_' among the column names, returns the position, and we we that to extract that names


e2.m <- reshape2::melt(e2[,c('agec','date',log_var_names)], id.vars=c('agec','date'))

```

Plot the controls

```{r}
p4 <- ggplot(e2.m, aes(x=date, y=value, group=variable, col=variable)) +
      geom_line()  +
      ylab("Death by cause") +  #add y axis label
      xlab("Date") +   #add x axis label
      theme_classic() +  #choose a 'theme'
      geom_vline(xintercept=as.Date('2006-01-01'), col='gray', lty=2) + #add reference line for vaccine introduction
      facet_wrap(~agec, scales='free_y') + #different plot for each age group, each with unique scale (scale='free')
      theme(panel.spacing= unit(2,'lines') , axis.text.x=element_text(angle=90))

p4
```

That's it. Now we have a mostly clean dataset that we are ready to analyze in the next session!
















